---
title: "Course Project 2 - PM2.5"
author: "Davide"
date: "Thursday, August 21, 2014"
output: html_document
---

Set up the environment:
```{r}
rm(list = ls())
setwd('C:/Nuova cartella/Exploratory Data Analysis/Course Project 2 - PM2.5')
```

Read the two data sets into dataframe objects:
```{r}
NEI = readRDS("summarySCC_PM25.rds")
str(NEI)
head(NEI)

SCC = readRDS("Source_Classification_Code.rds")
str(SCC)
head(SCC)
```

## Question #1
*Have total emissions from PM2.5 decreased in the United States from 1999 to 2008?*

Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008

```{r}
# Totalize Emissions per each year
total_emissions_per_year <- with(NEI,  tapply(Emissions, year, sum, na.dm = TRUE))
str(total_emissions_per_year)
total_emissions_per_year
names(total_emissions_per_year)

# Get labels for plot labelling
years_range = range(NEI$year)
str(years_range)
years_range_character = as.character(years_range)

main <- paste('Total US PM 2.5 Emissions\n', years_range_character[1], 'to', years_range_character[2], sep = ' ')
main

# Plot total emissions vs year
plot(names(total_emissions_per_year), total_emissions_per_year, col = 'red', lwd = 2, type = 'l', xlab = 'Year', ylab = 'Emissions [tons]', main = main, xaxt = 'n')
axis(side = 1, at = names(total_emissions_per_year), labels = names(total_emissions_per_year))
points(names(total_emissions_per_year), total_emissions_per_year, pch = 19, col = 'blue')
```

## Question #2
*Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008?*

Use the base plotting system to make a plot answering this question.
```{r}
NEI_baltimore <- subset(NEI, NEI$fips == '24510')

total_emissions_per_year_baltimore <- with(NEI_baltimore,  tapply(Emissions, year, sum, na.dm = TRUE))

str(total_emissions_per_year_baltimore)
total_emissions_per_year_baltimore
names(total_emissions_per_year_baltimore)

years_range_baltimore = range(NEI_baltimore$year)
str(years_range_baltimore)
years_range__baltimore_character = as.character(years_range_baltimore)

main <- paste('PM 2.5 Emissions\n', 'Baltimore City, MD,', years_range__baltimore_character[1], 'to', years_range__baltimore_character[2], sep = ' ')
main

plot(names(total_emissions_per_year_baltimore), total_emissions_per_year_baltimore, col = 'red', lwd = 2, type = 'l', xlab = 'Year', ylab = 'Emissions [tons]', main = main, xaxt = 'n')
axis(side = 1, at = names(total_emissions_per_year_baltimore), labels = names(total_emissions_per_year_baltimore))
points(names(total_emissions_per_year_baltimore), total_emissions_per_year_baltimore, pch = 19, col = 'blue')
```

## Question #3
*Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008?*

Use the ggplot2 plotting system to make a plot answer this question.
```{r}
#install.packages('ggplot2')

require('ggplot2')
library(ggplot2)
```

```{r}
NEI_baltimore_type <- split(NEI_baltimore, NEI_baltimore$type)
str(NEI_baltimore_type)
summary(NEI_baltimore_type)

names <- c()
df <- NULL
for (i in 1:length(NEI_baltimore_type)) {

  name <- unique(NEI_baltimore_type[[i]]$type)
  name <- paste('Emissions', name, sep = '.')
  name <- gsub('-', '_', name)
  names <- c(names, name)
  
  v1 <- with(NEI_baltimore_type[[i]], tapply(Emissions, year, sum, na.dm = TRUE))
  years <- names(v1)
  v1 <- unname(v1)
  
  types <- rep(name, times = 4)
  
  print(years)
  print(v1)
  print(types)
  
  df1 <- data.frame(year = years, Emissions = v1, type = types)
  print(df)
  print(df1)
  if (is.null(df))
    df = df1
  else
    df <- rbind(df, df1)
  
}
```

```{r}
Palette1 <- c('red','green','blue','violet','black')

ggplot(df, aes(year, Emissions, colour = type)) +
geom_point() + 
geom_point(aes(size = Emissions)) + 
scale_colour_manual(values = Palette1) +
geom_line(aes(group = factor(df$type))) +
theme(legend.position = 'bottom', plot.title = element_text(lineheight = 1.6, face = 'bold', size = 22)) +
ylab('Emissions [ton]') +
ylab('Year') +
ggtitle('PM 2.5 Emissions by type, 1999 to 2008\nBaltimore City, MD')
```

## Question #4
*Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?*

The strategy I'll use to extract the emissions of coal related sources is ad  follows:

* Using a regular expression (`.*[cC]oal.*`) we'll match the values of SCC's EI.Sector columns that are coal-related. The reg exp means:
  * ".*"    -- matches any number ("*") of any characters ("."). Read it as "any character (.) as many times (*) as you'd like"
  * "[cC]"  -- matches either the lowercase "c" or uppercase "C" just once
  * "oal"   -- matches exactly that series of characters
  * ".*"    -- matches any character.
  
* the corresponding values of SCC are extracted;
* these values are then compared with NEI's SCC columns and matching rows are selected.

Let's see this in action:
```{r}
coal_related_EI.Sectors <- unique(grep('.*[cC]oal.*', SCC$EI.Sector, ignore.case = TRUE, value = TRUE))
coal_related <- SCC$EI.Sector %in% coal_related_EI.Sectors
length(coal_related)
head(coal_related)
sum(coal_related)
```

Only `r sum(coal_related)` rows of SCC dataset (`r sum(coal_related)/dim(SCC)[1]` %) are coal-related. Now get the corresponding SCC codes, and use them to subset the NEI dataset:
```{r}
coal_related_SCC <- SCC[coal_related, ]$SCC
NEI_coal_related <- NEI[NEI$SCC %in% coal_related_SCC, ]
head(NEI_coal_related)
dim(NEI_coal_related)
```

Here we have a new data set od emissions consisting of `r dim(NEI_coal_related)[1]` observations (`r dim(NEI_coal_related)[1]/dim(NEI)[1]` %).

A little bit of consistency check is worthwile:
```{r}
sum(unique(NEI_coal_related$SCC) %in% coal_related_SCC)
```

```{r}
total_emissions_per_year <- with(NEI_coal_related,  tapply(Emissions, year, sum, na.dm = TRUE))
total_emissions_per_year
```

```{r}
years_range = range(NEI$year)
str(years_range)
years_range_character = as.character(years_range)

main <- paste('US PM 2.5 Emissions\n', 'Coal-related, ', years_range_character[1], 'to', years_range_character[2], sep = ' ')
main

# Plot total emissions vs year
plot(names(total_emissions_per_year), total_emissions_per_year, col = 'red', lwd = 2, type = 'l', xlab = 'Year', ylab = 'Emissions [tons]', main = main, xaxt = 'n')
axis(side = 1, at = names(total_emissions_per_year), labels = names(total_emissions_per_year))
points(names(total_emissions_per_year), total_emissions_per_year, pch = 19, col = 'blue')
```

## Question #5
*How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?*

I'll use the solving procedure of Question #4, changing the regular expression to `.*[vV]ehicles*`.

```{r}
vehicles_related_EI.Sectors <- unique(grep('.*[vV]ehicles*', SCC$EI.Sector, ignore.case = TRUE, value = TRUE))
vehicles_related <- SCC$EI.Sector %in% vehicles_related_EI.Sectors
vehicles_related_SCC <- SCC[vehicles_related, ]$SCC
NEI_vehicles_related <- NEI[NEI$SCC %in% vehicles_related_SCC, ]
```

Question ask for vehicles emissions for Baltimore City, MD only, so we have an additional subsetting step to perform:
```{r}
NEI_vehicles_related_baltimore <- NEI[NEI_vehicles_related$fips == '24510', ]
dim(NEI_vehicles_related_baltimore)
```

Now we can proceed with as before:
```{r}
total_emissions_per_year <- with(NEI_vehicles_related_baltimore,  tapply(Emissions, year, sum, na.dm = TRUE))
total_emissions_per_year
years_range = range(NEI$year)
str(years_range)
years_range_character = as.character(years_range)

main <- paste('Baltimore City, MD PM 2.5 Emissions\n', 'Vehicle-related, ', years_range_character[1], 'to', years_range_character[2], sep = ' ')
main

# Plot total emissions vs year
plot(names(total_emissions_per_year), total_emissions_per_year, col = 'red', lwd = 2, type = 'l', xlab = 'Year', ylab = 'Emissions [tons]', main = main, xaxt = 'n')
axis(side = 1, at = names(total_emissions_per_year), labels = names(total_emissions_per_year))
points(names(total_emissions_per_year), total_emissions_per_year, pch = 19, col = 'blue')
```

## Question #6
*Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?*

```{r}
NEI_vehicles_related_losangeles <- NEI[NEI_vehicles_related$fips == '06037', ]

total_emissions_baltimore <- with(NEI_vehicles_related_baltimore,  tapply(Emissions, year, sum, na.dm = TRUE))
total_emissions_losangeles <- with(NEI_vehicles_related_losangeles,  tapply(Emissions, year, sum, na.dm = TRUE))

years_range = range(NEI_vehicles_related_baltimore$year)
years_range_character = as.character(years_range)

emissions_range = range(total_emissions_baltimore, total_emissions_losangeles, na.rm = TRUE)

main <- paste('PM 2.5 Emissions\nBaltimore City, MD vs Los Angeles County\n', 'Vehicle-related, ', years_range_character[1], 'to', years_range_character[2], sep = ' ')

plot(names(total_emissions_baltimore), total_emissions_baltimore, col = 'red', lwd = 2, type = 'l', xlab = 'Year', ylab = 'Emissions [tons]', main = main, xaxt = 'n', ylim = emissions_range)
points(names(total_emissions_baltimore), total_emissions_baltimore, pch = 19, col = 'red')

lines(names(total_emissions_losangeles), total_emissions_losangeles, pch = 19, col = 'blue', lwd =2, type = 'l')
points(names(total_emissions_losangeles), total_emissions_losangeles, pch = 19, col = 'blue')
axis(side = 1, at = names(total_emissions_baltimore), labels = names(total_emissions_baltimore))

legend('topleft', c('Baltimore City, MD', 'Los Angeles County'), col = c('red', 'blue'), pch = 19)
```
